Pyg4ometry: a Python library for the creation of Monte Carlo radiation transport physical geometries

### 2020-08-31 public/paper/pyg4ometry.tex
Paper update

%% main text
\section{Introduction} \label{sec:introduction}
There are numerous different software codes to simulate the passage of particles through material, radiation transport (RT) programmes, these 
include MCNP \cite{Mcnp_Werner}, FLUKA \cite{Fluka_Ferrari,Fluka_Bohlen}, Geant3 \cite{Geant3_Brun} and Geant4 \cite{Geant4_Agostinelli}. 
All these codes are based on the Monte Carlo technique but each code either has a particular specialism, simulation methodology or target user community.  
RT Monte Carlo (RTMC) simulations have diverse uses, shielding calculations for radiological protection, detector performance, medical imaging and therapy and space radiation environment  simulation are some examples. A fundamental requirement of all of the codes is to supply a computer readable description of the 
physical geometry that  the particles are passing through.  The creation of geometry files is typically a very time consuming activity and the simulation validity 
and performance is directly dependent on the quality of the geometry. The requirement of a geometry system that allows the conversion between files prepared 
for different codes will enable cross checks of the physics processes in different particle transport codes. Parallel to the creation of geometry files for RT 
programs, usually computer assisted design (CAD) files exist for systems which need to be simulated in RT programmes. The fundamental geometric 
representations in CAD files is typically not amenable for RT programmes.  For these reasons it is advantageous to create a software tool that allows 
particle transport code users to rapidly develop geometry files, convert between common RT programme geometry formats and finally import CAD models.

@@

\begin{figure*}[hbt]
  \normalsize
  \centering
  \includegraphics[width=0.7\textwidth]{./diagrams/workflow.pdf}
  \caption{\label{fig:workflow}Pyg4ometry geometry conversion workflow.}
\end{figure*}

@@

\subsection{Visualisation} \label{sec:visualisation}
A  geometry hierarchy can be viewed using the popular Visualisation Toolkit (VTK). No separate scene graph is required as the Geant4 volume hierarchy is sufficient 
to place the meshes associated with each physical volume. A daughter volume is placed within a logical volume with a rotation $\mathbf{R}$, scale $\mathbf{S}$ and 
translation $\mathbf{T}$.

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=5cm]{./diagrams/lvToPv.pdf}
\caption{Placement of physical volume inside a logical volume}
\label{fig:lvToPv}
\end{center}
\end{figure}

\begin{eqnarray}
\mathbf{M}	  	& = & \mathbf{M}_m \mathbf{M}_d  = \mathbf{S}_m \mathbf{R}_m  \mathbf{S}_d \mathbf{R}_d\\
\mathbf{T}	 		& = & \mathbf{M}_m \mathbf{T}_d + \mathbf{T}_m= \mathbf{S}_m \mathbf{R}_m \mathbf{T}_d + \mathbf{T}_m,
\end{eqnarray}
where the subscript $p$ indicates parent volume and $d$ indicates daughter volume.

The python version of Geant4's physical volume is also used to store visualisation attributes like the solids 
color, surface or wireframe representation and visibility. Overlaps detected in the discretised mesh geometry can be displayed separately to allow a user to remove
the potential overlaps. This is dependent on the accuracy of the mesh generated and is a particular problem with meshes based on curved surfaces.

%%%%%%%%%%%%%%%%% Adding properties to volumes
A pure geometry description is an insufficient description for a RT simulation. Often other attributes need to be assigned for example material definition, magnetic field or  
optical surface properties.

\subsection{Overlap detection}
All RTM Ctransport codes can not handle spatial overlap between two geometric objects and likely are to have ill defined behaviour when tracking particles  
through such a situation.  A key feature of pyg4ometry is the detection of potential overlaps efficiently, it does this by performing an intersection operation etween solid instances and determining if the resulting mesh is empty. Figure~\ref{fig:overlap} shows the potential overlaps, a) protrusion of a daughter from the mother, b) and intersection with finite volume and c) an overlap where two daughters share a face. 
\begin{figure}[htbp]
\begin{center}
\includegraphics[width=8cm]{./diagrams/overlap.pdf}
\caption{Schematic of the three different types of overlap between daughters of a mother logical volume.}
\label{fig:overlap}
\end{center}
\end{figure}

@@

@@

\section{Rapid geometry modelling}
Given the python scripting interface, expression  and tessellation engines. It is possible for a user to rapidly specify the geometrical layout of the RT problem, vary 
the parameters of the geometry and visualise the geometry. Finally when the desired result without geometry overlaps is finalised a GDML file can be output
from the internal memory representation.

@@

@@

\section{GDML to FLUKA conversion}
It is relatively straight forward to convert Geant4 geometry to FLUKA. Each of the Geant4 solids can be mapped to a FLUKA \verb|REGION|
Some solids in Geant4 directly map to a FLUKA body, others need a construction of a simple FLUKA CSG tree. Table~\ref{tab:geant2fluka}
lists the mapping between geant4 solids and the bodies used to compose a fluka region.

@@

There are some important issues which must be considered to perform an accurate conversion. 
A Geant4 logical volume solid might need to be created from a fluka region composing of many zones, 
so $R_1 = +z_1 | +z_2$, similarly a void is required to locate daughter volume placements which again 
might also be converted to a region of multiple zones $R_2= +z_3| z_4$. So creating the fluka region is then 
\begin{eqnarray}
R_1 - R_2 	& = & (+z_1 | +z_2) - ( +z_3 | +z_4) 			\\
			& = & (+z_1 - z_3 - z_4)  | (+z_2 - z_3 - z_4)	
\label{eqn:setDiff}
\end{eqnarray}
Geant4 has the boolean solids associated with difference, union and intersection, so in addition to Equation~\ref{eqn:setDiff}, both $R_1 \cup R_2$ and 
$R_1 \cap R_2$ is required in fluka notation, so 
\begin{eqnarray}
R_1 \cup R_2 	& = & (+z_1 | +z_2)  \cup ( +z_3 | +z_4) \\
			& = & +z_1 | +z_2 |  +z_3 | +z_4
\label{eqn:setUnion}
\end{eqnarray}
and 
\begin{eqnarray}
R_1 \cap R_2 	& = & (+z_1 | +z_2) \cap ( +z_3 | +z_4) \\
			& = & +z_1 +z_3  | +z_1 +z_4 | +z_2 +z_3 | +z_2 +z_4
\label{eqn:setIntersection}
\end{eqnarray}

@@

\subsection{Non-Convex solid decomposition}
In general the base solids used by Geant4 allow non-convex solids, eg Polycone, Xtru, the Twisted.  These particular solids are problematic when 
converting geometry to a FLUKA readable format, as non-convex solids can only be created by union of convex zones. 
There are two groups which need to considered, firstly those where 2D polygonal section needs to be decomposed these include Polycone, 
Polyhedron and Xtru, secondly those where three dimensional convex decomposition is required  TwistedBox, TwistedTrap, TwistedTrd, 
TwistedTube and Tessellated. These solids are converted to CGAL 
Nef polyhedra \colorbox{yellow}{XXX CITE} and decomposed to convex polyhedra using a \colorbox{yellow}{XXX} algorithm.

\subsection{Disjunctive normal form and degenerate surfaces}
Typically FLUKA will decompose a \verb|REGION| into disjunctive normal form (DNF), this normal form is characterised as the union of intersections, so 
\begin{equation}
R = z_1 \; | \;z_2\;  | \; z_3 | \; z_4 \dots
\end{equation}
Defining \verb|REGION|s in terms of DNF allows the rapid test if a point is inside or outside. Testing each zone $z_i$ of $R$ can terminate if a point 
is determined to be inside any of convex zones. In general Equation~\ref{eqn:logicalSubtraction} is not in a DNF. If there are many levels of logical-physical 
volume placement, then recursive application of  Equation~\ref{eqn:logicalSubtraction} will create a nested set of parentheses. There are some conditions 
where a general boolean expression can yield an exponential explosion of the final DNF. There are well known algorithms to convert logical expressions to 
DNF  \colorbox{yellow}{XXX}. Pyg4ometery can simplify parentheses from a \verb|REGION| by creating a sympy \colorbox{yellow}{XXX CITE} expression and 
using its logical module. Before version 4.0 FLUKA would simplify all regions to DNF and some converted geometry would generate too many zones and would fail.
FLUKA version 4 allows control of the DNF expansion.

@@

\begin{enumerate}
\item Degenerate body look up 
\item Bodes with/without transformation 
\item Non-convex solids, convex decomposition (TwistedBox, TwistedTrap, TwistedTrd, TwistedTube, Tessellated)
\item Scaled solid, Placement scale 
\item Replica, Division, Parameterised
\item Recursive application of \ref{eqn:setDiff}, \ref{eqn:setUnion}, \ref{eqn:setIntersection}, \label{eqn:logicalSubtraction} speed up using meshes
\end{enumerate}

\section{CAD/STL to GDML conversion }
STEP and IGES files can be loaded into pyg4ometry, via an interface based on FreeCAD. FreeCAD is an open source CAD/CAM program, which in turn 
is based on OpenCASCADE. A STEP CAD file could be considered as a hierarchical tree of {\it part assemblies} and {\it part features}, where a part assembly is a 
collection of part features. A part feature can be used to create  a triangular mesh which can used to create a pyg4ometry tessellated solid. Assignment of materials 
and visualisation attributes must be performed by the user after conversion to GDML as rarely will CAD/CAM packages included the detailed information required 
for RT codes.

\begin{lstlisting}[caption={A simple pyg4ometry python script to load a STEP file},label={lst:pythonCADLoading}, language=Python]
    reader = pyg4ometry.freecad.Reader("FileName.step")
    reader.relabelModel()
    reader.convertFlat()
    logical = reader.getRegistry().getWorldVolume()
\end{lstlisting}

\begin{figure*}[htbp]
\begin{center}
\includegraphics[width=1.0\textwidth]{./model-scene/SectorBend.pdf}
\caption{Example conversion of a simple FLUKA geometry to GDML, left panel :  
the original model in FreeCAD, right panel : the GDML geometry 
viewed in pyg4ometry. The example is a sector bend dipole electromagnet.}
\label{fig:cad-gdml}
\end{center}
\end{figure*}

\section{Complete simulation example}
Pyg4ometry is designed to be as flexible as possible and offer the user a wide range of usage styles and workflows. The following sections describe a selection of 
possible applications.

\begin{enumerate}
\item Merging Registries 
\item Removing volumes 
\item Editing solid
\item Converting to assembly solids 
\end{enumerate}

\begin{figure*}[htbp]
\begin{center}
\includegraphics[width=1.0\textwidth]{./model-scene/model.pdf}
\caption{Complete compositing example using pyg4ometry.}
\label{fig:model}
\end{center}
\end{figure*}

@@

\begin{enumerate}
\item Graphical front end viewer
\item C++ output
\item CAD BREP to FLUKA converter (a la McCAD) 
\item Integration of other tools (DAGMC) 
\item Far from perfect conversions
\end{enumerate}

>>>>

%% main text
\section{Introduction} \label{sec:introduction}
There are numerous different software codes to simulate the passage of particles through material, radiation transport (RT) programmes, these 
include MCNP \cite{Mcnp_Werner}, FLUKA \cite{Fluka_Ferrari,Fluka_Bohlen}, Geant3 \cite{Geant3_Brun} and Geant4 \cite{Geant4_Agostinelli}. 
All these codes are based on the Monte Carlo technique but each code either has a particular specialism, simulation methodology or target user community.  
RT Monte Carlo (RTMC) simulations have diverse uses, shielding calculations for radiological protection, detector performance, medical imaging and therapy and space radiation environment  simulation are some examples. A fundamental requirement of all of the codes is to supply a computer readable description of the 
physical geometry that  the particles are passing through.  The creation of geometry files is typically a very time consuming activity and the simulation validity 
and performance is directly dependent on the quality of the geometry. The requirement of a geometry system that allows the conversion between files prepared 
for different codes will enable cross checks of the physics processes in different particle transport codes. Parallel to the creation of geometry files for RT 
programs, usually computer assisted design (CAD) files exist for systems which need to be simulated in RT programmes. The fundamental geometric 
representations in CAD files is typically not amenable for RT programmes.  For these reasons it is advantageous to create a software tool that allows 
particle transport code users to rapidly develop geometry files, convert between common RT programme geometry formats and finally import CAD models. 
\colorbox{yellow}{More explanation of Geant4 and FLUKA}

@@

\begin{figure*}[hbt!]
  \normalsize
  \centering
  \includegraphics[width=0.7\textwidth]{./diagrams/workflow.pdf}
  \caption{\label{fig:workflow}Pyg4ometry geometry conversion workflow.}
\end{figure*}

@@

\subsection{Visualisation} \label{sec:visualisation}
A  geometry hierarchy can be viewed using the popular Visualisation Toolkit (VTK). No separate scene graph is required as the Geant4 volume hierarchy is sufficient 
to place the meshes associated with each physical volume. A daughter volume is placed within a logical volume with a rotation $\mathbf{R}_d$, scale $\mathbf{S}_d$ and translation $\mathbf{T}_d$.

\begin{figure}[htb!]
\begin{center}
\includegraphics[width=5cm]{./diagrams/lvToPv.pdf}
\caption{Placement of physical volume inside a logical volume}
\label{fig:lvToPv}
\end{center}
\end{figure}

The transformation  $\mathbf{M}$  and translation $\mathbf{T}$ from mother to daugher is 
\begin{eqnarray}
\mathbf{M} 	& = &  \mathbf{S}_d  \mathbf{R}_d \\
\mathbf{T} 	& = &  \mathbf{T}_d
\end{eqnarray}

If the mother volume is placed in the world then the placement transformation $\mathbf{M}_w$ and translation $\mathbf{T}_w$ 
\begin{eqnarray}
\mathbf{M}_w	  	& = & \mathbf{M}_m \mathbf{M}_d  = \mathbf{S}_m \mathbf{R}_m  \mathbf{S}_d \mathbf{R}_d				\label{eqn:worldToDaughter1}\\
\mathbf{T}	_w 		& = & \mathbf{M}_m \mathbf{T}_d + \mathbf{T}_m= \mathbf{S}_m \mathbf{R}_m \mathbf{T}_d + \mathbf{T}_m  \label{eqn:worldToDaughter2}
\end{eqnarray}
where the subscript $p$ indicates parent volume and $d$ indicates daughter volume. Given a hierarchy of logical and physical volumes, 
Equations~\ref{eqn:worldToDaughter1} and \ref{eqn:worldToDaughter2} can be used recursively depending on the number of nested volumes.

The python version of Geant4 physical volume is also used to store visualisation attributes like the solid's 
colour, surface or wireframe representation and visibility. Overlaps detected in the mesh geometry are
stored in the Logical Volume and can be  displayed separately to allow a user to understand the reason for 
the overlaps.

%%%%%%%%%%%%%%%%% Adding properties to volumes
Geometry alone is an insufficient information for a RTMC simulation. Often other attributes need to be assigned to regions of space for example material definition, magnetic field or optical properties. These physical properties can be used to define the visualisation attributes of a volume.

\subsection{Overlap detection}
All RTMC transport codes can not handle spatial overlap between two geometric objects and likely are to have ill defined behaviour when tracking particles  
through such a situation.  A key feature of pyg4ometry is the detection of potential overlaps efficiently, it does this by performing an intersection operation etween solid instances and determining if the resulting mesh is empty. Figure~\ref{fig:overlap} shows the potential overlaps, a) protrusion of a daughter from the mother, b) and intersection with finite volume and c) an overlap where two daughters share a face. 
\begin{figure}[htbp]
\begin{center}
\includegraphics[width=8cm]{./diagrams/overlap.pdf}
\caption{Schematic of the three different types of overlap between daughters of a mother logical volume.}
\label{fig:overlap}
\end{center}
\end{figure}

@@

The accuracy of the overlap visualisation is dependent on the accuracy of the meshes generated, for example problems might occur 
with meshes based on curved surfaces.

@@

\section{Rapid geometry modelling}
Given the python scripting interface, expression  and tessellation engines. It is possible for a user to rapidly specify the geometrical layout of the RT problem, vary 
the parameters of the geometry and visualise the geometry. Finally when the desired result without geometry overlaps is finalised a GDML file can be output
from the internal memory representation. Some example code is presented in Listing~\ref{lst:pythonRapidModelling}. The structure should be familiar to regular 
users of Geant4, first a \verb|Registry| need to be created to store all the pyg4ometry objects; followed by constants;  then materials and properties, solids and logical and physical volumes; finally the whole geomtery can be saved as a GDML file and/or visualised using VTK.

@@

The pyg4ometry python code in the example is approximately as expressive 
as the GDML it writes. The benefit of wrapping GDML/Geant4 in Python allows 
to very rapid prototyping of geometry without concerns of compilation of C++ (in the 
case of Geant4) or writing well formed XML (in the case of GDML). Effectively
the Python interpreter checks input for syntax errors when using pyg4ometry classes. 
Another key benefit is the ability to use the pyg4ometry code to create 
programatic converters between different geometry languages.

@@

\section{GDML to FLUKA conversion}
It is relatively straight forward to convert Geant4 geometry to FLUKA. Each of the Geant4 
solids can be mapped to a FLUKA \verb|REGION|. Some solids in Geant4 directly map to 
a FLUKA body, others need a construction of a simple FLUKA CSG tree. Table~\ref{tab:geant2fluka}
lists the mapping between geant4 solids and the bodies used to compose a fluka region.

@@

There are some important issues which must be considered to perform an accurate conversion. 
A Geant4 logical volume solid might need to be created from a fluka region composing of many zones, 
so $R_1 = +z_1 | +z_2$, similarly a void is required to locate daughter volume placements which again 
might also be converted to a region of multiple zones $R_2= +z_3| z_4$. So creating the fluka region is then 
\begin{eqnarray}
R_1 - R_2 	& = & (+z_1 | +z_2) - ( +z_3 | +z_4) 			\\
			& = & (+z_1 - z_3 - z_4) \;  | \; (+z_2 - z_3 - z_4)	
\label{eqn:setDiff}
\end{eqnarray}
Geant4 has the boolean solids associated with difference, union and intersection, so in addition to 
Equation~\ref{eqn:setDiff}, both $R_1 \cup R_2$ and $R_1 \cap R_2$ is required in fluka notation, so 
\begin{eqnarray}
R_1 \cup R_2 	& = & (+z_1 | +z_2)  \cup ( +z_3 | +z_4) \\
			& = & +z_1 | +z_2 |  +z_3 | +z_4
\label{eqn:setUnion}
\end{eqnarray}
and 
\begin{eqnarray}
R_1 \cap R_2 	& = & (+z_1 | +z_2) \cap ( +z_3 | +z_4) \\
			& = & +z_1 +z_3  \; | +z_1 +z_4 \; | +z_2 +z_3 \; | +z_2 +z_4
\label{eqn:setIntersection}
\end{eqnarray}

@@

\subsection{Non-Convex solid decomposition}
In general the base solids used by Geant4 allow non-convex solids, eg Polycone, Xtru, the Twisted.  These 
particular solids are problematic when converting geometry to a FLUKA readable format, as non-convex solids 
can only be created by union of convex zones. There are two groups which need to considered, firstly those 
where 2D polygonal section needs to be decomposed these include Polycone, Polyhedron and Xtru, secondly 
those where three dimensional convex decomposition is required  TwistedBox, TwistedTrap, TwistedTrd, 
TwistedTube and Tessellated. The second classification of non convex solids are converted to CGAL 
Nef polyhedra \colorbox{yellow}{XXX CITE} and decomposed to convex polyhedra using a 
\colorbox{yellow}{XXX} algorithm.

\subsection{Disjunctive normal form and degenerate surfaces}
Typically FLUKA will decompose a \verb|REGION| into disjunctive normal form (DNF), this normal form is 
characterised as the union of intersections, so 
\begin{equation}
R = z_1 \; | \;z_2\;  | \; z_3 	\; | \; z_4 \dots
\end{equation}
Defining \verb|REGION|s in terms of DNF allows the rapid test if a point is inside or outside. Testing each zone 
$z_i$ of $R$ can terminate if a point is determined to be inside any of convex zones. In general 
Equation~\ref{eqn:logicalSubtraction} is not in a DNF. If there are many levels of logical-physical volume 
placement, then recursive application of  Equation~\ref{eqn:logicalSubtraction} will create a nested set of 
parentheses. There are some conditions where a general boolean expression can yield an exponential 
explosion of the final DNF. There are well known algorithms to convert logical expressions to DNF 
\colorbox{yellow}{XXX}. Pyg4ometery can simplify parentheses from a \verb|REGION| by creating a 
sympy \cite{10.7717/peerj-cs.103} expression and using its \verb|logical| module. Before version 4.0 FLUKA 
would simplify all regions to DNF and some converted geometry would generate too many zones and would fail.
FLUKA version 4 allows control of the DNF expansion.

@@

\begin{enumerate}
\item Degenerate body look up 
\item Bodes with/without transformation 
\item Non-convex solids, convex decomposition (TwistedBox, TwistedTrap, TwistedTrd, TwistedTube, Tessellated)
\item Scaled solid, Placement scale 
\item Replica, Division, Parameterised
\item Recursive application of \ref{eqn:setDiff}, \ref{eqn:setUnion}, \ref{eqn:setIntersection}, \ref{eqn:logicalSubtraction} speed up using meshes
\end{enumerate}

\section{CAD/STL to GDML conversion }
STEP and IGES files can be loaded into pyg4ometry, via an interface based on FreeCAD. FreeCAD is an 
open source CAD/CAM program, which in turn is based on OpenCASCADE. A STEP CAD file could be 
considered as a hierarchical tree of {\it part assemblies} and {\it part features}, where a part assembly is a 
collection of part features. A part feature can be used to create  a triangular mesh which can used to create 
a pyg4ometry tessellated solid. Assignment of materials and visualisation attributes must be performed by 
the user after conversion to GDML as rarely will CAD/CAM packages included the detailed information required 
for RT codes.
\begin{lstlisting}[caption={A simple pyg4ometry python script to load a STEP file},label={lst:pythonCADLoading}, language=Python]
    reader = pyg4ometry.freecad.Reader("FileName.step")
    reader.relabelModel()
    reader.convertFlat()
    logical = reader.getRegistry().getWorldVolume()
\end{lstlisting}

\begin{figure*}[htb!]
\begin{center}
\includegraphics[width=1.0\textwidth]{./model-scene/SectorBend.pdf}
\caption{Example conversion of a simple FLUKA geometry to GDML, left panel :  
the original model in FreeCAD, right panel : the GDML geometry 
viewed in pyg4ometry. The example is a sector bend dipole electromagnet.}
\label{fig:cad-gdml}
\end{center}
\end{figure*}

\section{Complete simulation example}
Pyg4ometry is designed to be as flexible as possible and offer the user a wide range of usage styles 
and workflows. The following sections describe a selection of 
possible applications.
%
\begin{enumerate}
\item Merging Registries 
\item Removing volumes 
\item Editing solid
\item Converting to assembly solids 
\end{enumerate}
%
\begin{figure*}[htb!]
\begin{center}
\includegraphics[width=1.0\textwidth]{./model-scene/model.pdf}
\caption{Complete compositing example using pyg4ometry.}
\label{fig:model}
\end{center}
\end{figure*}

@@

\begin{enumerate}
\item Graphical front end viewer
\item C++ output
\item CAD BREP to FLUKA converter (a la McCAD) 
\item Integration of other tools (DAGMC) 
\item Far from perfect conversions
\item Export mesh for other applications (GPU optics)
\item Convergence FLUKA/Geant4 - Moira 
\item Replace G4 polyhedron CSG
\end{enumerate}
